Concept tests for statistics
============================


Distribution theory
-------------------

### Problem: Exponential distribution

This problem concerns customers in a coffee shop.   Assume that we can model waiting times as an Exponential($\lambda$) distribution.   Under this model, the expected waiting time is 12 minutes.

It is now 12:20pm.   In which of the following scenarios do you expect to get served the soonest

- (a) The last customer arrived at 12:10
- (b) The last customer arrived at 12:15
- (c ) The last customer arrived at 12:00
- (d) The last customer arrived at 12:08

You may pick none, some or all.


### Comments

This problem aims to engage students around the memoryless property of the exponential distribution.   It is simple to prove that the Exponential has a memoryless problem, reasonably straightforward to show that the Exponential is the only continuous distribution family to have this property.   It may be harder to really appreciate what this means.   As a point of detail it may be necessary to explain whether $\lambda$ denotes waiting time or the reciprocal of waiting time.

Exploratory Data Analysis
-------------------------

### Problem: Measures of location

In which histogram is the median larger than the mean

```{r, echo = FALSE, error=FALSE,results='hide',message=FALSE}
require(sn)
par(mfrow = c(2,2))
hist(rnorm(10000), xlim = c(-10,10), freq = FALSE, col = "red", main = "a", xlab = "x")
hist(c(rnorm(999), rep(8,10)), col = "red", xlim = c(-10,10), freq = FALSE, main = "b", xlab = "x")
hist(rsn(1000,0,3,5), xlim = c(-10,10), freq = FALSE, col = "red", main= "c", xlab = "x")
hist(rsn(1000,0,3,-5), xlim = c(-10,10), freq = FALSE, col = "red", main =  "d", xlab = "x")
```

### Comments

There are quite a few suggestions like this floating around in various books and journals.   The idea is to try to get students to understand what the histograms are telling them, rather than providing the jargon "left skew" etc.   In sport statistics for example, students often know what a right tailed distribution tells them (some players get a lot of goals, most players get a few)


Random variables and notation
-----------------------------

### Problem: Notation

The variables $Y_1, Y_2, \ldots, Y_n$ are independently and identically distributed as Normal random varables with parameters $\mu$ and $\sigma_2$.   We can write $Y_i \sim N(\mu, \sigma^2)$ where $i = 1, 2, \ldots, n$.

Now consider the random variable $\bar{Y} = \frac{Y_1 + Y_2 + \cdots + Y_n}{n}$.   What can you say about the following variances:

- (a) $Var(Y_i)$ increases with $n$
- (b) $Var(Y_i)$ decreases with $n$
- (c ) $Var(Y_i)$ is unchanged with $n$
- (d) $Var(\bar{Y})$ increases with $n$
- (e) $Var(\bar{Y})$ decreases with $n$
- (f) $Var(\bar{Y})$ is unchanged with $n$


### Comments

The question has a little to do with the concepts that can be hidden beneath notation.   It could be extended by asking about the variance of $y_i$ and $\hat{y}$


```{r}
blah <- function(x){
  inv_count <- 0
  n <- length(x)
  
  for(i in c(1:n)){
    for(j in (i:n)){
      if(x[i] > x[j])
        inv_count <- inv_count + 1
 }}
  return (inv_count)
}
 
  numbers = c(1, 20, 6, 4, 5)
numbers <- c(2, 4, 1, 3, 5 )
numbers <- as.vector(read.csv("IntegerArray.txt", header = FALSE)[,1])
blah(numbers)
blah(x)
  printf(" Number of inversions are %d \n", getInvCount(arr, 5));
  getchar();
  return 0;
}



int  _mergeSort(int arr[], int temp[], int left, int right);
int merge(int arr[], int temp[], int left, int mid, int right);
  
/* This function sorts the input array and returns the
   number of inversions in the array */

  
  mergeSort <- function(x){
    n <- length(x)
    return _mergeSort(x, temp, 0, n - 1);
}
  

 DomergeSort <- function(x, temp, left, right){
  if (right > left){
    mid = (right + left)/2;
  
    /* Inversion count will be sum of inversions in left-part, right-part
      and number of inversions in merging */
    inv_count  = DomergeSort(x, temp, left, mid);
    inv_count += DomergeSort(x, temp, mid+1, right);
  
    /*Merge the two parts*/
    inv_count += merge(arr, temp, left, mid+1, right);
  }
  return inv_count;
}
  
/* This funt merges two sorted arrays and returns inversion count in
   the arrays.*/
int merge(int arr[], int temp[], int left, int mid, int right)
{
  int i, j, k;
  int inv_count = 0;
  
  i = left; /* i is index for left subarray*/
  j = mid;  /* i is index for right subarray*/
  k = left; /* i is index for resultant merged subarray*/
  while ((i <= mid - 1) && (j <= right))
  {
    if (arr[i] <= arr[j])
    {
      temp[k++] = arr[i++];
    }
    else
    {
      temp[k++] = arr[j++];
  
     /*this is tricky -- see above explanation/diagram for merge()*/
      inv_count = inv_count + (mid - i);
    }
  }
  
  /* Copy the remaining elements of left subarray
   (if there are any) to temp*/
  while (i <= mid - 1)
    temp[k++] = arr[i++];
  
  /* Copy the remaining elements of right subarray
   (if there are any) to temp*/
  while (j <= right)
    temp[k++] = arr[j++];
  
  /*Copy back the merged elements to original array*/
  for (i=left; i <= right; i++)
    arr[i] = temp[i];
  
  return inv_count;
}
  
/* Driver progra to test above functions */
int main(int argv, char** args)
{
  int arr[] = {1, 20, 6, 4, 5};
  printf(" Number of inversions are %d \n", mergeSort(arr, 5));
  getchar();
  return 0;
}




```
